Composition-climate paper

### 2016-12-03 Response_to_reviewers.Rmd

---
bibliography: goringetal_references.bib
csl: ecology.csl
output:
  word_document:
    reference_docx: word_template.docx  
  pdf_document:
    pandoc_args: -V geometry:vmargin=1in -V geometry:hmargin=1in
  md_document:
    variant: markdown_github
pandoc_args: -H margins.sty
---

# Response to Reviewers

Dear Dr. Thrall:

We submit for your consideration the revised manuscript "*Effect of historical land-use and climate change on tree-climate relationships in the upper Midwestern United States.*"

We have carefully revised the manuscript in response to reviewer comments.  In these revisions, we have paid particular attention to the potential limitations raised by the reviewers regarding the 1) comparing the PLS and FIA data and 2) choice of climate normal.  We completed several new analyses to test the sensitivity of our results to choice of climate normal and created a separate subsection within the Discussion assessing the robustness of our findings.
<!-- be sure to fill in the word count -->
We have also edited the manuscript for clarity and to respond to the other points raised by the reviewers.  We include below a point-by-point response to all reviews.  The revised version has XX  words and 6 figures and tables.  The main text of revised manuscript is now slightly beyond the 5,000 word limit, so please advise whether we should trim further to meet the target. We have expanded the Supplementary Information to show several additional analyses.

@@

> The manuscript written by Goring and Williams (“Effect of historic land-use and climate change on tree-climate relationships in the northern United States”; ELE-00978-2016) addresses an important topic in global change research: what is the relative influence of climate vs. land-use change on contemporary species distributions. The question is relevant as the degree to which the two drivers are both important indicates potentially major errors in future projections previously published using species distribution models (SDMs). The authors present quite a clever method for addressing the issue: i.e., building SDMs with both historical and contemporary vegetation and climate data. This paper provides novel perspective concerning how our understanding of the role of climate change in vegetation change is still limited. The paper is exceedingly well written and the conclusions seem well-supported by the methods and results.

We thank the reviewer for the positive feedback and constructive criticisms.  We address specific points below.

> My primary critique (which I admit is relatively minor) is that I found the logic revolving around the “confounding effects” and how the results relate to that a bit difficult. In particular, there likely needs to be a bit of clarification around the concept and a justification for why treating two non-significant effects and a set of mixed effects (one significant and one not so) as equivalent within this framework.

To address this point, we’ve expanded Table 1 to indicate the individual terms (dV and dL) and their significance.  This lets the reader see for themselves what goes into 'confounding'.  We also added to the Introduction (4th paragraph) the sentence “*These confounding effects may hinder detection of climatic forcing of historical species distributions.*”  As it turns out, in the primary analyses presented here, all instances of confounding involve a significant change in climate and no significant change in vegetation (Table 1)

@@

> Line 143-144: Shouldn’t miles be replaced with SI? I realize that the miles are in reference to the historical context of the PLS, but it might be good to keep things consistent throughout.

We have changed to *km*, retaining the reported miles in parentheses.

> Lines 174-176: This is a key point, but expanding a bit would be helpful. For example, did the paper in review examine issues of detectability? I realize this is addressed in the discussion, but a bit more detail would be good here.

This Goring et al. paper is now accepted and in press.  We have added the following paragraph:

*Given the differences in sampling design, FIA data should be better at detecting species presence at the level of individual plots, while detectability at the level of 8km grid cell (the analysis unit of this study) should be similar between FIA and PLS data or higher in the latter.  Within grid cells, there is a tradeoff between the intensive sampling within FIA plots versus more PLS points per grid cell and more uniform sampling of landscape heterogeneity.  The PLS data may be better at detecting tree presence in low-density regions, due to the variable radius of PLS plots.  Both the FIA and PLS datasets both have good coverage in the upper Midwest, with 5,172 8km grid cells with at least one FIA plot and 7,376 grid cells with at least one PLS point.*

And have added a section on 'Potential Limitations' to the Discussion.

@@

> Line 207: Here and in a few other places, d_l is used, but later d_v is used. Please make symbology consistent throughout the paper.

We have changed all references from $d_{l}$ to $d_{v}$.

> Lines 217-218: I found the labeling of lack of significance in “either or both changes” as confounding to be hard to connect with the definition provided at Line 93. For example, Fagus shows a strong overall change in distribution relative to precipitation, but is classes as confounding. Presumably, this is a situation where climate was important, but history was not. I suppose a bit more clarification regarding what is meant by confounding and why grouping taxa where neither climate nor history has an effect with taxa where only one has an effect is necessary.

We've changed the table to make things more clear.  'Confounding' is now restricted to situations when there is significant effects of climate change on the tree distributions but no significant effect of land use.

@@

> Lines 231-232: Looking at the maps, it seems that this statement is limited by the fact that the current study does not incorporate the northern range limit for many species. Can the authors address this limitation at some point?

We have added the paragraph:  *For all species, the range within the upper Midwest incompletely samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.* to address this point.

@@

> Lines 337-351: The authors discuss agriculture, but not timber harvesting and changes in fire regime in terms of the land-use drivers. It seems that the latter two are potentially important for some of the regional patterns. Additionally, I wonder if some of this could be discussed by looking at gains and losses relative to NLCD land-use layers. Perhaps that was in the supplement and I missed it.

We have added the sentence: *While logging and altered fire regimes have altered community composition in the north, they have not widely excluded taxa from the northern regions and associated climates, except for * Tsuga * (`r fig_nums("gain_loss", display="cite")`), a shade-tolerant and late-successional taxon.*

@@

We have changed the x axis to a log10 scale and added more tick marks to the x-axis to make it clear that it is non-linear.

![](figures/hellingerplot.png)

> Table 1: Some of the taxa are listed with their scientific names and others by common names. Replace common names with scientific names

We have corrected the names so that only scientific names are displayed.

@@

## Referee: 2

> This paper compares tree/climate relationships between two time periods and concludes that modern forest data sets may not be adequate for quantifying the realized niches of tree species because tree populations may not be in equilibrium with environment under rapid land use and climate change and the data sets have limitations in estimating tree presence and absence.  This is a very important point and the paper will likely be of very high interest.  This comparison of pre-settlement times with modern times is unique in the species distribution modeling literature and nicely demonstrates the key points.

> The paper is well written and the graphics are generally very good.

We thank Reviewer 2 for the positive feedback and constructive criticism.  We have performed several new analyses and revised existing analyses in response.  We describe these further below.

@@

> I think the paper does not adequately point out the limitations of the data sets used for the analyses and that the authors overstate the likely differences in realized niches between the historic and current time periods.

> The paper would be strengthened by stating more strongly the limitations of the methods and that the results may be over estimates and that data are not currently available to more accurately estimate the effect.

> Moreover, other data sets should be called upon to try to corroborate the range shifts and changes in climate-tree relationships that emerge from the analyses.  Also the ecological mechanisms that might underlay the effects of land use and climate on tree species distributions should be more fully discussed.

We thank the reviewer for these comments. We now more fully address the limitations through multiple solutions. 1) We have added references to long term weather records and used their early 19th century estimates as the basis for sensitivity analysis (see below).  2) We have experimented with two modern climate normals, representing time of sampling and time of establishment. 3) We strengthened the literature reiew section that indicates prior support for widespread land use change, and its impacts on tree species distributions in the region. 4) Lastly, in the Discussion we have reitterated what we believe are some of the limitations of the current study.

@@

### More specific comments

> 2  "north-central" US would be more accurate? Also change this throughout the ms.

Here and elsewhere, we have adopted ‘upper Midwestern US’ throughout ms. as the most spatially precise geographic place name.

@@

> 175-176 While that study found little effect of the differences between the PLS and FIA data sets, be sure to cover fully the potential effect of these differences on results of this study.  One has to presume that the differences in realized niches between historic and present would be substantially smaller if FIA was sampling low density and widespread occurrence of species as did the PLS.

Please see our response above to Reviewer 1 and the expanded treatment of this question in the Methods and Discussion.

@@

> 181-182 Provide evidence that these climate variables have been shown to be strong correlates of tree species presence in previous species distribution modeling studies.  It is important to do your analysis on climate variables that are known to be ecologically relevant to tree niche dimensions.

We have added the sentences: *These climate variables represent three major climatic controls on plant distributions: summer warmth, winter minimum temperatures, and moisture availability; these or similar bioclimatic variables are widely used in plant distributional modeling [@woodward1987climate;@iverson1998predicting;@maiorano2013building].*

@@

> 183-185   First, it is a serious stretch to use 1895-1919 climate to represent tree/climate relationships during 1830-1910.  The verbal statement (without statistical analysis) that the dendro-based PDSI did not change over that period is not very convincing.  Moreover, the analysis specifically should be done on the climate predictors used in the niche modeling.  Also, evaluate to what extent does this PDSI index represent trends in Tmin and the other predictors used?

We respectfully disagree that it is a serious stretch to use 1895-1915 climates to represent 1830-1910 climates.  The PDSI data shows no evidence of long-term trend between 1800 and 1910, based on a LOESS line fitted to the time series shown in Fig. 1 (see other responses below).

To further address this point, we have added a review of early meteorological records from US military forts. These records do indicate some warming in the region (Baker et al, 1985, Burnett et al. 2010).   We have added the following text to the first paragraph of the results.

*Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al. 1985) and Fort Leavenworth, Kansas (Burnette et al. 2010). The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1). Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).*

We have used these findings as the basis of a simple sensitivity analysis to the Supplementary Information (Supplementary Figure 3) in which adjust the historic climate data and historic-modern differences based on 19th-century trends observed in the US fort data.  This analysis is summarized by the following new sentence in the results:

*These patterns are unchanged by sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).*

@@

> Another issue is lack of discussion about the time period when the forests sampled by FIA got established. It is typical in SDM studies to test the strength of models from different time periods in tree SDM work and select the time period that is most strongly associated with species presence.  In the western US, we found 1950-1980 to produce stronger models than 1980-2010.

We have addressed this point by adding an analysis in which we use climate normals for 1950-1980 for the modern climate instead of 1990-2014.  We report the results from the latter in the main paper, and have added several new figures to the supplementary information.  We’ve also analyzed tree age in the FIA data, which shows a mean age of 52, 56, and 49 years in WI, MN, and MI.  Based on this, the 1950-1980 is a good alternative time window for the time of tree establishment.
We also note that this paper is not intended as an exercise in creating SDMs, a point we will return to later, but take the point of the association between vegetation and climate varies between establishment and long term survival of forest stocks.
We’ve added the following text to the manuscript:

*The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.  The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here. The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region. We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.*

@@

> A third issue is the role of climate variability over the 1895 to 2014 time period.  Decadal oscillations are common in this period in many parts of the US.  Trees are likely integrating over 2 or more of these climate cycles.  SDMs should be done to best represent the climate and variability that selected for tree presence when the trees were measured.  Your analysis would be strengthened with statistical analysis of climate patters from 1895 to present for your predictor variables and an analysis of which time period produces the best SDMs.

We believe that this point is now addressed in responses to other comments. We have added a LOESS curve to Figure 1 (in response to a later  comment), we have selected an alternate modern climate normal that represents the period of establishment, and have provided additional support to the choice of the modern normal.   Note too that we specifically chose 30-year climatologies in order to smooth out the effects of interannual and interdecadal-scale variability on our results.

@@

> 195 It is important to discuss why univariate climate tree presence functions are relevant to the climate factors that strongly influence tree species presence or absence.  SDM modeling typically shows that a particular combination of climate and habitat factors is much more strongly associate with tree presence and any random climate variable.  Your analysis would be stronger to do model selection for best multivariate SDM functions and then compare these functions between historic and present.

We apreciate this point and thank the reviewer for bringing it up. Please note that this analysis is not intended to focus on the construction of SDMs, but rather to show the joint effects of historical land use and climate change on tree-climate relationships. We respectfully believe that a multivariate SDM would obscure these shifting relationships by blending together the climate variables. Further, a multi-model SDM approach might weight climate variables differently across models, through time.  We believe that there is value and insight in focusing on individual climate variables and the distribution of trees along individual axes.

Shifts in the strength of SDMs between time scales and potential utility of climate variables in SDM construction is worthwhile, but a separate analysis and paper.

@@

> 211-218 The analyses should control for spatial autocorrelation, simple t-tests would seem to have an inflated sample size due not including spatial autocorrelation in the model.

We have now included spatial autocorrelation in this analysis, and describe the methods in the Supplementary Information.

@@

> 226-230 and Figure 1.  Statistical analysis should be done of the climate trends, showing a least a trend line and a 95% confidence interval.  A moving window regression might better show breakpoints in the trend line.  This is needed because the data show high interannual variability and it is not obvious by eyeing these graphics for which variables trends exist or not. Also, say in the legend what the red dots are.

We have added a LOESS curve onto Figure 1, representing a span of 20 years, along with the SE fit. We have added light shading over boxes corresponding to the chosen climate normals. As indicated above, we have provided a more detailed analysis of the choice of normal periods and described the effect of the choice on outcomes.  During the figure revisions, we removed the red dots.

@@

> 231-239 I find the maps of range loss in Fig 3 to be stunningly large.  I am not sure anything in the literature would document such large losses in tree species presence from land use or 100 yrs of climate change.  It would seem important to use other data sets to validate these estimates of range loss.  Are the patterns in these maps real or mostly due to the differences among the PLS and FIA data sets?  If real, what are the mechanisms of local extinction?

These losses are well supported in the literature. Rhemtulla et al (2007) cite estimates of land use conversion up to 79% for southern deciduous savanna, and Rhemtulla et al (2009) show that losses in oak-hickory forest complexes surpass 5,000,000 ha.  As noted in the new sentences: "*Rhemtulla et al. [-@rhemtulla2009historical] estimate that Pinus strobus and Tsuga canadensis now occupy only 4% of their original coverage in Wisconsin. Wildland ecosystems have been converted to agricultural and urban land use [@rhemtulla2009historical], particularly in historical prairie and savanna ecosystems (Figure 1f), where 79% of deciduous savanna was converted to cropland [@rhemtulla2007regional].*"

@@

> 347-350 Is it accurate to infer that ag has “largely eliminated open forests” which does not sample low density forests? It seems likely that most of these tree species are present in at least low densities in the areas where your maps show they are not present.

As above, we believe that these patterns are real.  We have added text to the Methods and Discussion about  differences in the FIA and PLS data and potential effects on detecting low-density trees.

@@

> Lines 361-364.  The topic of this sentence should be elaborated upon to more fully point out the limitations of the study and appropriate inference. Say how much the estimates of forest loss and niche shifts may be overestimated.

Please see the response to the next comment.

> 365-370 These sentences are a bit of bait and switch.  Yes, readers will agree that SDM based on FIA may not fully represent the realized niche of tree species.  However, some of your argument hinges on the magnitude of the error that is made by using FIA.  I think your paper way over states that error.  The paper would have more impact if you were careful to emphasize the limitations of the data sets for this analysis and say that the degree to which FIA leads to SDM error is not knowable with available data sets.

We have created a standalone section in the Discussion (and the various sensitivity analyses described above) that more fully works through these points.  Our takeaway message here is that the limitations are real and need considering, but do not detract from the main finding that historic land use and climate changes in this region have been large and have interacted to alter tree-climate relationships.

@@

## Referee 3

> Goring and Williams investigated the overall distribution changes and relative influence of historic land use and climate change on these changes in the upper Midwestern United States. This is an interesting topic because climate change has been changing tree species distribution, meanwhile land use change may be interacting with climate change to compound, confound, or contract these changes. However, the manuscript is not very compelling for several reasons:

We thank the reviewer for these comments and hope that the revised manuscript addresses these concerns.

@@

> Firstly, its rigor and clarity do not meet the standards for Ecology Letters. For example,

> Title: The authors used the upper Midwestern United States throughout the manuscript. But why the authors used the northern United States in the title?

See response to Reviewer 2.  We have changed the title to ‘upper Midwestern United States’.

@@

> This study equally emphasized the distribution changes for 15 common tree genera over the last two centuries and the effects of climate change and land use change on these changes. Was it better to say “the effects of historical land use and climate change on species distribution changes” in the tittle?

We respectfully believe the title should stand as is. This paper does indeed present evidence for both changes in geographic distributions and tree-climate relationships.  However, we believe the latter is of more interest to the Ecology Letters readership and have focused the paper’s attention on it.

@@

> Abstract: P2-L34-37, forest inventory data cannot be used to project forest responses to climate change. Rather, forest inventory data were used in models to make projections. It was distribution-climate relationships. In my opinion, this study was not about the stability and representativeness of distribution-climate. This study was to build up or extend the distribution-climate relationships.

We appreciate the clarification of the reviewer regarding the first point and have addressed it by changing the introductory sentence to:

@@

*Contemporary forest inventory data are widely used to understand environmental controls on tree species distributions and to construct models to project forest responses to climate change, but the stability and representativeness of contemporary tree-climate distributions is poorly understood*

We respectfully disagree with the reviewer regarding the second point.  We believe that the central aspect of this paper is its analysis of the stability of tree-climate relationships over the past two centuries.  The paper shows that, for many taxa, land use following EuroAmerican settlement in the north-central US limits the representitiveness of modern inventory-based distribution models.   We do not attempt here to use historical and contemporary data to build better predictions for the future, but have inserted a pointer to other papers where that has been done:

*One solution is to use both historical and contemporary data when building species distribution models (Nogues-Bravo et al. 2016 Maiorano et al. 2013).*

@@

> 3) One of the objectives in this manuscript was to investigate the effects of climate change, land use change, and pathogen outbreaks on distribution changes (L125-126). However, the pathogen outbreaks were nearly ignored in the title and methods, results, and discussion sections.

### 2016-12-03 Composition_Ecology.Rmd
Revised response to reviews, minor edits to main text

In mapping tree distributions in the upper Midwest we consider three classes: "gain," where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss," where a tree was present in a grid cell during the PLS but absent in the FIA; and "continuous presence," where a tree was present within a grid cell in both FIA and PLS eras.  We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA observations (Supplementary Table  2). For all species, the range within the upper Midwest incompletely samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.

@@

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternatives for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data. We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplementary Information. There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatiotemporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].

@@

Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C between the 1895-1924 and 1985-2014 normals (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C (see Supplementary Information for estimates based on the 1950 - 1979 normals). $P_{ann}$ increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.  Long term observer and military fort observations suggest warming during the 19^th^ century, with increases in mean annual temperature at Fort Snelling, Minnesota [@baker1985minnesota] and Fort Leavenworth, Kansas [@burnette2010daily]. The latter shows faster increases during winter (`r round(clim_change[4], 1) + 0.5`^o^C) than summer(`r round(clim_change[2], 1) + 0.35`^o^C). Hence, the PRISM-based estimates of historical climate change may be too conservative for $T_{min}$ and $T_{diff}$, and too large for $T_{max}$.

@@

Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all p < 0.0001).  The directionality of shifts is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ and $P_{ann}$ are higher (`r fig_nums("boxplot", display = "cite")`).  *Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while *Larix* has the smallest. Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm). These patterns are unchanged by sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).

>>>>

In mapping tree distributions in the upper Midwest we consider three classes: "gain," where a tree species was present within a grid cell in the FIA, but absent in the PLS; "loss," where a tree was present in a grid cell during the PLS but absent in the FIA; and "presence," where a tree was present within a grid cell in both FIA and PLS eras.  We present gain and loss estimates for all grid cells (Supplementary Table 1) and for the subset of grid cells with both PLS and FIA observations (Supplementary Table  2). For all species, the range within the upper Midwest incompletely samples species ranges, however, the region contains several major ecotones and the southern and western range limit for many tree species [@curtis1959vegetation] so this region comprises an important boundary in geographic and climate space.

@@

Three periods were selected for calculation of climate normals: 1895 to 1919 for the historical period and two alternatives for the modern period; 1950 - 1979 representing the average period of establishment for stand in the upper Midwest and 1985 - 2014 covering the period of sampling for the FIA data. We report results for the 1985 - 2014 normals, with the 1950-1980 results shown in Supplementary Information. There is an unavoidable temporal mismatch between the historical climate data and the PLS survey period, on the order of 50 to 70 years; few meteorological records exist in this region prior to 1895 CE, and no gridded dataset exists at high spatiotemporal resolution for this time period for all climate variables.  The effects of temporal mismatch should be minor given that Northern Hemisphere temperature trends were generally small during the 19^th^ century [@mann2009global].  To check whether an 1895 to 1919 CE climate normal is a reasonable surrogate for pre-settlement climates we compare the instrumental record against dendroclimatic reconstructions of the Palmer Drought Severity Index [@heim2002review] (`r fig_nums("land_use_clim", display = "cite")`) and to early temperature observations from US military forts and observer stations, which extend to the 1830s [@baker1985minnesota;@burnette2010daily].
<!-- Re-inserted paragraph below (I think it's important to explain why we picked the two modern normals that we did.  Note that there is a missing ref below, highlighted in bold.) -->
The two alternative modern climate normals correspond to time of sampling versus time of tree establishment.  The 1990-2014 normal encompasses the period of sampling (2000 to 2015) for the FIA data used here. The 1950-1979 normal encompasses the time of establishment of FIA trees, which have an average age of 50 years in this region (**REF**). We report results for the 1990-2014 normals, with the 1950-1979 results shown in Supplementary Information.

@@

Regionally, $T_{min}$ increased `r round(clim_change[4], 1)`^o^C between the 1895-1924 and 1985-2014 normals (`r fig_nums("land_use_clim", display = "cite")`), $T_{max}$ changed by `r round(clim_change[2], 1)`^o^C, while $T_{diff}$ changed by `r round(clim_change[3], 1)`^o^C (see Supplementary Information for estimates based on the 1950 - 1979 normals). $P_{ann}$ increased by `r round(clim_change[1], 1)`mm.  The Palmer Drought Severity Index is correlated to $P_{ann}$ trends from 1895 onward (r = 0.43) and, with LOESS regression, indicates no major climatic trends between 1800 and 1900 (`r fig_nums("land_use_clim", display = "cite")`e), suggesting that early 20^th^-century precipitation is a reasonable proxy for mid-19^th^-century precipitation.  Early temperature observations from US military forts indicate warming between the middle 19^th^ and early 20^th^ centuries at Fort Snelling, Minnesota (Baker et al. 1985) and Fort Leavenworth, Kansas (Burnette et al. 2010). The latter shows faster increases during winter (0.10oC decade-1) than summer (0.07oC decade-1). Hence, our estimates of historic climate change may be too conservative for Tmin and Tdiff and too large for Tmax (Supplementary Information).

@@

Shifts in tree-climate distributions are significant for all genera and climate variables (`r fig_nums("boxplot", display = "cite")`), even accounting for multiple comparisons and spatial autocorrelation (all p < 0.0001).  The directionality of shifts is consistent across genera: *e.g.*, modern $T_{max}$ is lower for all genera, while $T_{min}$ and $P_{ann}$ are higher (`r fig_nums("boxplot", display = "cite")`).  *Ulmus* shows the largest increase in minimum temperature (`r fig_nums("boxplot", display = "cite")`; $\delta_{Tmin}$ = 4.2 ^o^C), while *Larix* has the smallest. Maximum annual temperature declines most for *Ostrya*/*Carpinus* ($\delta_{Tmax}$ = -1.3 ^o^C) and least for *Ulmus* ($\delta_{Tmax}$ = -0.4 ^o^C).  *Ulmus* shows the largest increase in mean precipitation ($\delta_{Pann}$ = 86mm), and *Larix* the smallest ($\delta_{Pann}$ = 42mm). These patterns are unchanged by a sensitivity analysis in which the historical temperature changes are augmented by estimated temperature trends reported from the historical records, but the $T_{max}$ changes are sensitive to choice of modern climate normal (Supplemental Information).
